---
title: "Computational Text Analysis Assignment"
author: "Molly Smith"
date: "2026-04-14"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

## Download relevant packages
```{r}
library(kableExtra)
library(tidyverse) 
library(readr)
library(stringr)
library(tidytext)
library(quanteda)
library(textdata)
library(dplyr)
library("ggplot2")
library("plyr")
library("gdata")
library("data.table")
```

## Download the data
```{r}
data = fread("/Users/mollysmith/Downloads/dataverse_files-2/uk_data.csv", encoding="UTF-8")


data$date = as.Date(data$date)


# Table 2: Examples: Emotive and neutral speeches
example1 = subset(data, id_speech==854597)
example1$emotive_rhetoric
example1$text

example2 = subset(data, id_speech==778143)
example2$emotive_rhetoric
example2$text

# Create time variable
data$time= NA
data$time[data$date>=as.Date("2001-01-01") & data$date<=as.Date("2001-06-30")] = "01/1"
data$time[data$date>=as.Date("2001-07-01") & data$date<=as.Date("2001-12-31")] = "01/2"
data$time[data$date>=as.Date("2002-01-01") & data$date<=as.Date("2002-06-30")] = "02/1"
data$time[data$date>=as.Date("2002-07-01") & data$date<=as.Date("2002-12-31")] = "02/2"
data$time[data$date>=as.Date("2003-01-01") & data$date<=as.Date("2003-06-30")] = "03/1"
data$time[data$date>=as.Date("2003-07-01") & data$date<=as.Date("2003-12-31")] = "03/2"
data$time[data$date>=as.Date("2004-01-01") & data$date<=as.Date("2004-06-30")] = "04/1"
data$time[data$date>=as.Date("2004-07-01") & data$date<=as.Date("2004-12-31")] = "04/2"
data$time[data$date>=as.Date("2005-01-01") & data$date<=as.Date("2005-06-30")] = "05/1"
data$time[data$date>=as.Date("2005-07-01") & data$date<=as.Date("2005-12-31")] = "05/2"
data$time[data$date>=as.Date("2006-01-01") & data$date<=as.Date("2006-06-30")] = "06/1"
data$time[data$date>=as.Date("2006-07-01") & data$date<=as.Date("2006-12-31")] = "06/2"
data$time[data$date>=as.Date("2007-01-01") & data$date<=as.Date("2007-06-30")] = "07/1"
data$time[data$date>=as.Date("2007-07-01") & data$date<=as.Date("2007-12-31")] = "07/2"
data$time[data$date>=as.Date("2008-01-01") & data$date<=as.Date("2008-06-30")] = "08/1"
data$time[data$date>=as.Date("2008-07-01") & data$date<=as.Date("2008-12-31")] = "08/2"
data$time[data$date>=as.Date("2009-01-01") & data$date<=as.Date("2009-06-30")] = "09/1"
data$time[data$date>=as.Date("2009-07-01") & data$date<=as.Date("2009-12-31")] = "09/2"
data$time[data$date>=as.Date("2010-01-01") & data$date<=as.Date("2010-06-30")] = "10/1"
data$time[data$date>=as.Date("2010-07-01") & data$date<=as.Date("2010-12-31")] = "10/2"
data$time[data$date>=as.Date("2011-01-01") & data$date<=as.Date("2011-06-30")] = "11/1"
data$time[data$date>=as.Date("2011-07-01") & data$date<=as.Date("2011-12-31")] = "11/2"
data$time[data$date>=as.Date("2012-01-01") & data$date<=as.Date("2012-06-30")] = "12/1"
data$time[data$date>=as.Date("2012-07-01") & data$date<=as.Date("2012-12-31")] = "12/2"
data$time[data$date>=as.Date("2013-01-01") & data$date<=as.Date("2013-06-30")] = "13/1"
data$time[data$date>=as.Date("2013-07-01") & data$date<=as.Date("2013-12-31")] = "13/2"
data$time[data$date>=as.Date("2014-01-01") & data$date<=as.Date("2014-06-30")] = "14/1"
data$time[data$date>=as.Date("2014-07-01") & data$date<=as.Date("2014-12-31")] = "14/2"
data$time[data$date>=as.Date("2015-01-01") & data$date<=as.Date("2015-06-30")] = "15/1"
data$time[data$date>=as.Date("2015-07-01") & data$date<=as.Date("2015-12-31")] = "15/2"
data$time[data$date>=as.Date("2016-01-01") & data$date<=as.Date("2016-06-30")] = "16/1"
data$time[data$date>=as.Date("2016-07-01") & data$date<=as.Date("2016-12-31")] = "16/2"
data$time[data$date>=as.Date("2017-01-01") & data$date<=as.Date("2017-06-30")] = "17/1"
data$time[data$date>=as.Date("2017-07-01") & data$date<=as.Date("2017-12-31")] = "17/2"
data$time[data$date>=as.Date("2018-01-01") & data$date<=as.Date("2018-06-30")] = "18/1"
data$time[data$date>=as.Date("2018-07-01") & data$date<=as.Date("2018-12-31")] = "18/2"
data$time[data$date>=as.Date("2019-01-01") & data$date<=as.Date("2019-06-30")] = "19/1"
data$time[data$date>=as.Date("2019-07-01") & data$date<=as.Date("2019-12-31")] = "19/2"

data$time2 = data$time
data$time2 = str_replace(data$time2, "/", "_")

data$stage = 0
data$stage[data$m_questions==1]= 1
data$stage[data$u_questions==1]= 2
data$stage[data$queen_debate_others==1]= 3
data$stage[data$queen_debate_day1==1]= 4
data$stage[data$pm_questions==1]= 5
```

## Create year and speech-text variables and filter EU speeches
```{r}
eu_speeches <- data %>% # Data saved as eu_speeches
  dplyr::mutate(
    year = as.integer(format(date, "%Y")), # New year variable created
    speech_text = text # New variable created for text
  ) %>%
  dplyr::filter( # Filters speeches from 2012 to 2016
    year >= 2012,
    year <= 2016,
    stringr::str_detect(speech_text, stringr::regex("\\bEU\\b|European Union", ignore_case = TRUE))
  ) %>% # Filters speeches to keep only speeches that mention EU or European Unions
  dplyr::select(id_mp, date, year, party, words, speech_text) # Keeps only selected variables
```

## Check that the date conversion worked:
```{r}
eu_speeches %>% # Use eu_speeches dataset
  dplyr::count(year)  # Count how many rows/speeches fall into each year value
```

## Create a yearly dataset of EU speeches
```{r}
eu_speeches_yearly <- eu_speeches %>% # Start with the eu_speeches dataset and assign the result to a new object called eu_speeches_yearly
  dplyr::group_by(year) %>% # Group the data by year
  dplyr::summarise( # Summarise each yearly group into one row per year
    speech_text = paste(speech_text, collapse = " "), # Combine all speech text from the same year into one long text string
    total_words = sum(words, na.rm = TRUE), # Add together the word counts for all speeches in each year, ignoring missing values
    n_speeches = dplyr::n(), # Count how many speeches there are in each year
    .groups = "drop" # Remove the grouping after summarising so the result is no longer grouped
  )
```

## Check the year distribution in the full dataset
```{r}
data %>% # Use the original dataset 
  dplyr::mutate(year = as.integer(format(date, "%Y"))) %>%  # Take the date variable, extract just the year part, and store it as a numeric year variable
  dplyr::count(year, sort = TRUE)   # Show the frequency of each year, ordered by the most common year first
```

## Count speeches matching the EU keyword filter
```{r}
data %>%
  dplyr::mutate(year = as.integer(format(date, "%Y"))) %>% # Extract the year from each date and store it as an integer variable called year
  dplyr::filter(year >= 2012, year <= 2016) %>% # Restrict the dataset to observations from 2012 to 2016
  dplyr::summarise( # Summarise the filtered observations into one result
    matches = sum(stringr::str_detect(text, stringr::regex("\\bEU\\b|European Union", ignore_case = TRUE)), na.rm = TRUE) # Count the number of rows where the text contains 'EU' or 'European Union', regardless of case
  )
```

## Refine the corpus using more specific EU terms
```{r}
eu_pattern <- stringr::regex( # Create a regular expression to search for EU-related terms
  "\\bEU\\b|\\bEuropean Union\\b|\\bBrexit\\b|\\bsingle market\\b|\\bcustoms union\\b|\\bfree movement\\b|\\bBrussels\\b|\\bArticle 50\\b|\\bEuropean Commission\\b|\\breferendum\\b", # Define the list of keywords and phrases to match in the text
  ignore_case = TRUE  # Make the search case-insensitive 
)

eu_speeches <- data %>%  # Use the original dataset as the staring point
  dplyr::mutate( # Add new variables needed
    year = as.integer(format(date, "%Y")), # Extract the year from the date variable
    speech_text = text # Create a new variable called speech_text
  ) %>%
  dplyr::filter( # Keep only rows that meet all of the following conditions
    year >= 2012, # Keep speeches from 2012 onwards
    year <= 2016, # Keep speeches up to and including 2016
    stringr::str_detect(speech_text, eu_pattern) # Keep only speeches whose text contains at least one of the EU-related terms in eu_pattern
  ) %>%
  dplyr::select(id_mp, date, year, party, words, speech_text) # Keep only the variables needed for the final eu_speeches dataset
```

## Inspect the filtered dataset
```{r}
head(eu_speeches) # Display the first few rows of eu_speeches
```

## Preprocess and tokenise the data
```{r}
tidy_data <- eu_speeches %>% # Create a new object called tidy_data
  dplyr::filter(party %in% c("Labour", "Conservative")) %>% # Keep only these parties
  dplyr::select(id_mp, date, year, party, speech_text) %>% # Keep relevant columns
  dplyr::mutate(
    speech_text = tolower(speech_text), # Lowercase everything
    speech_text = stringr::str_replace_all(speech_text, "\\beuropean union\\b", "european_union"),
    speech_text = stringr::str_replace_all(speech_text, "\\bsingle market\\b", "single_market"),
    speech_text = stringr::str_replace_all(speech_text, "\\bcustoms union\\b", "customs_union"),
    speech_text = stringr::str_replace_all(speech_text, "\\bfree movement\\b", "free_movement")
  ) %>%
  tidytext::unnest_tokens(word, speech_text) %>% # Split text into tokens
  dplyr::filter(stringr::str_detect(word, "^[a-z_]+$")) %>% # Keep alphabetic words
  dplyr::anti_join(stop_words, by = "word") # Remove stop words
```

## Define the fear-related dictionary
```{r}
fear_dict <- c( # Create a dictionary vector
  "threat", "threats", "threatening", "risk", "risks", "risky", "danger", "dangerous", "dangers", "crisis", "uncertainty", "insecure",
  "insecurity", "fear", "fears", "feared", "anxiety",
   "warning", "warn", "warned", "vulnerable",
  "instability", "insecure", "terror", "terrorism",
  "extremism", "violence", "harm", "damage", "damages", "disaster", "disasters", "undermine", "unfair", "unfairly", "unfairness", "rigged", "deprived", "impoverished", "poverty", "exploiting", "unravelling", "straining", "fragile", "protect", "emergency", "homeless", "homelessness", "unfunded", "unworkable", "violent", "violence", "extremism", "evil", "desperate", "desperately", "austerity", "unemployment", "unaffordable", "attack", "difficult", "caution", "injury", "defeat", "defeated", "botched", "intrudes", "destabilising", "degraded", "cuts", "cut", "depression", "mess", "predator", "immoral", "burden", "burdens", "aggressive", "aggression", "instability", "unstable", "unsound"
)
```

## Match the dictionary to the corpus
```{r}
fear_matches <- tidy_data %>% # Use tokenised dataset
  filter(word %in% fear_dict) # Filter data so only dictionary matches are kept
```

## Inspect word frequencies
```{r}
fear_matches %>%  # Use fear_matches dataset
  dplyr::count(word, sort = TRUE) # Count and sort fear words 
```

## Compare fear-word counts by party
```{r}
fear_matches %>% 
  dplyr::count(party, sort = TRUE) # Count how many fear-word matches are associated with each party 
```

## Normalise by total word count
```{r}
total_words_by_party <- tidy_data %>% # Use tidy_data dataset
  dplyr::count(party, name = "total_words") # Count the total number of words associated with each party and store as total_words

fear_by_party <- fear_matches %>% # Use dataset containing only fear-related words
  dplyr::count(party, name = "fear_words") %>% # Count how many fear-related words are used by each party and store this as fear_words
  left_join(total_words_by_party, by = "party") %>% # Join the fear-word counts with the total word counts by matching on party
  mutate(fear_rate = fear_words / total_words * 1000) # Calculate the rate of fear words per 1,000 total words for each party

fear_by_party # Display the final table of fear-word counts, total words, and normalised fear rates by party
```

## Line graph of fear-related language by party, 2012-2016
```{r}
fear_by_year_party <- fear_matches %>% # Use fear-word matched dataset
  dplyr::filter(party %in% c("Labour", "Conservative")) %>%  # Keep only Labour and Conservative parties
  dplyr::group_by(year, party) %>% # Group the data by both year and party
  dplyr::summarise(fear_count = dplyr::n(), .groups = "drop") # Count the number of fear-related words in each year-party group, then remove the grouping

ggplot(fear_by_year_party, aes(x = year, y = fear_count, colour = party)) + # Create a line graph with year on the x-axis, fear_count on the y-axis, and separate colours for each party
  geom_line(linewidth = 1) + # Add lines to show the trend in fear-word counts over time
  geom_point(size = 2) + # Add points to show the exact value for each year
  labs(  # Add a title and axis/legend labels
    title = "Fear-related word counts in EU speeches by party, 2012–2016",  # Set the graph title
    x = "Year", # Label the x-axis
    y = "Number of fear-word matches", # Label the y-axis
    colour = "Party" # Label the legend
  ) +
  theme_minimal() # Apply a minimal theme
```

## Bar chart of fear-word counts by party, 2012-2016
```{r}
ggplot(fear_by_year_party, aes(x = factor(year), y = fear_count, fill = party)) + # Create bar chart
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  labs( 
    title = "Fear-related language in EU speeches by party, 2012–2016",
    x = "Year", 
    y = "Fear-word count", 
    fill = "Party" 
  ) +
  theme_minimal() 
```

## Bar chart of the proportion of fear-related words by party, 2012-2016
```{r}
fear_prop <- tidy_data %>% # Use tidy_data dataset
  dplyr::filter(party %in% c("Labour", "Conservative")) %>% # Keep only observations from the Labour and Conservative parties
  dplyr::group_by(year, party) %>% # Group the data by year and party so counts are calculated for each party in each year
  dplyr::summarise( # Summarise each year-party group into one row
    total_words = n(), # Count the total number of words in each year-party group
    fear_words = sum(word %in% fear_dict),  # Count how many of those words appear in the fear_dict dictionary
    fear_prop = fear_words / total_words, # Calculate the proportion of words that are fear-related
    .groups = "drop" # Remove grouping after summarising
  )

ggplot(fear_prop, aes(x = factor(year), y = fear_prop, fill = party)) +
  geom_col(position = "dodge") +
  labs( 
    title = "Proportion of fear-related words in EU speeches by party, 2012–2016", 
    x = "Year", 
    y = "Proportion of fear words",
    fill = "Party" 
  ) +
  theme_minimal() 
```

## Most frequent fear-related words by party:
```{r}
top_fear_words <- fear_matches %>% # Use dataset containing only fear-related word matches
  dplyr::filter(party %in% c("Labour", "Conservative")) %>%  # Keep only observations for the Labour and Conservative parties
  dplyr::count(party, word, sort = TRUE) %>% # Count how often each fear-related word appears for each party, sorting from most to least frequent
  dplyr::group_by(party) %>% # Group the data by party so the next step selects top words within each party separately
  slice_max(n, n = 10) %>%  # Keep the 10 most frequent fear-related words for each party
  ungroup() # Remove grouping so the result is a regular data frame

ggplot(top_fear_words, aes(x = reorder(word, n), y = n, fill = party)) + # Create a bar chart
  geom_col(show.legend = FALSE) +  # Draw bars using the count values
  facet_wrap(~ party, scales = "free") + # Create separate panels for each party
  coord_flip() +  # Flip the axes so the word labels are easier to read
  labs( # Add the title and axis labels
    title = "Most frequent fear-related words by party", # Set the main title of the chart
    x = "Word", # Label the x-axis
    y = "Count"  # Label the y-axis
  ) +
  theme_minimal() 
```

## Heatmap of fear-related word counts
```{r}
ggplot(fear_by_year_party, aes(x = factor(year), y = party, fill = fear_count)) + # Create a heatmap
  geom_tile() + # Add a title and axis/legend labels
  labs(
    title = "Heatmap of fear-related word counts", 
    x = "Year", 
    y = "Party", 
    fill = "Count" 
  ) +
  theme_minimal() 
```

## Word cloud fear-related words across both parties
```{r}
library(wordcloud) # Load the wordcloud package
library(RColorBrewer) # Load the RColorBrewer package

fear_words_party <- fear_matches %>% # Start with the dataset containing fear-related word matches
  dplyr::filter(party %in% c("Labour", "Conservative")) %>% # Keep only rows from the Labour and Conservative parties
  dplyr::count(word, sort = TRUE) # Count how often each fear-related word appears and sort from most  to least frequent

wordcloud( # Create a word cloud using the counted fear-related words
  words = fear_words_party$word, # Use the word column as the words to display
  freq = fear_words_party$n, # Use the count column as the frequency for each word
  min.freq = 2, # Only include words that appear at least 2 times
  max.words = 100, # Display up to 100 words in the word cloud
  random.order = FALSE, # Plot the most frequent words first rather than in random order
  scale = c(4, 0.8), # Set the size range of the words from largest to smallest
  colors = brewer.pal(8, "Dark2") # Apply an 8-colour palette from RColorBrewer
)
```

## Word cloud of Labour's most frequent fear-related words
```{r}
labour_cloud <- fear_matches %>% 
  dplyr::filter(party == "Labour") %>% # Keep only rows from the Labour party
  dplyr::count(word, sort = TRUE)

wordcloud( 
  words = labour_cloud$word, 
  freq = labour_cloud$n, 
  min.freq = 2, 
  max.words = 100, 
  random.order = FALSE,
  scale = c(4, 0.8), 
  colors = brewer.pal(8, "Blues") 
)
```

## Word count of Conservative's most frequent fear-related words
```{r}
conservative_cloud <- fear_matches %>% 
  dplyr::filter(party == "Conservative") %>% # Keep only rows from the Conservative party
  dplyr::count(word, sort = TRUE) 

wordcloud( 
  words = conservative_cloud$word, 
  freq = conservative_cloud$n, 
  min.freq = 2, 
  max.words = 100, 
  random.order = FALSE, 
  scale = c(4, 0.8), 
  colors = brewer.pal(8, "Reds") 
)
```

## Comparative word cloud of fear-related words by party
```{r}
library(tidyr) # Load tidyr package

comparison_data <- fear_matches %>% 
  dplyr::filter(party %in% c("Labour", "Conservative")) %>% # Keep only rows where the party is Labour or Conservative
  dplyr::count(party, word) %>% # Count how many times each word appears in each party
  pivot_wider(names_from = party, values_from = n, values_fill = 0) # Reshape the data so each party gets its own column, filling missing counts with 0

comparison_matrix <- as.matrix(comparison_data[, -1]) # Convert the party count columns into a matrix, excluding the first column (word)
rownames(comparison_matrix) <- comparison_data$word # Use the words as the row names of the matrix

comparison.cloud( # Create a comparison word cloud
  comparison_matrix, # Use the matrix of word counts for the comparison
  max.words = 100, # Show up to 100 words in the word cloud
  random.order = FALSE,# Plot words in a fixed order rather than random order
  colors = c("red", "blue") # Use red and blue as the colours for the two groups
)
```

## Create summary table of fear-words and proportions
```{r}
summary_table <- tidy_data %>% # 
  dplyr::filter(party %in% c("Labour", "Conservative")) %>% 
  dplyr::group_by(year, party) %>% # Group the data by year and party
  dplyr::summarise( # Create summary statistics for each year-party group
    total_words = n(), # Count the total number of words in each group
    fear_words = sum(word %in% fear_dict), # Count how many words in each group appear in the fear dictionary
    fear_proportion = fear_words / total_words, # Calculate the proportion of fear words out of all words
    .groups = "drop" # Remove grouping after summarising
  )

summary_table # Display in summary table
```

## Validation stage
## Check that the key variables exist and look correct
```{r}
names(eu_speeches) # Show the names of all columns in the eu_speeches dataset
str(data[, .(date, text, party, words, id_mp)]) # Display the structure of the selected columns from the data dataset
```

## Check that the EU filtering worked
```{r}
eu_speeches %>% # 
  dplyr::count(year) # Count how many rows there are for each year

nrow(eu_speeches) # Count the total number of rows in the eu_speeches dataset
```

## Inspect a sample of the filtered speeches
```{r}
eu_speeches %>% 
  dplyr::select(date, party, speech_text) %>% # Keep only the date, party, and speech_text columns
  head(10) # Show the first 10 rows
```

## Check party balance
```{r}
eu_speeches %>% 
  dplyr::count(party, sort = TRUE) # Count how many rows belong to each party, and sort from highest to lowest
```

## Check token count by party
```{r}
tidy_data %>% 
  dplyr::count(party, name = "token_count") # Count how many rows belong to each party and name the count column token_count
```

## Check that dictionary matches are sensible
```{r}
fear_matches %>% 
  dplyr::count(word, sort = TRUE) # Count how many times each word appears, and sort the results from most frequent to least frequent
```

## Check whether any key variables have missing values
```{r}
colSums(is.na(eu_speeches[, c("date", "party", "speech_text", "id_mp")])) # Count the number of missing (NA) values in the columns date, party, speech_text, and id_mp
```

## Check for blank or missing speech text
```{r}
eu_speeches %>% 
  summarise(blank_text = sum(speech_text == "" | is.na(speech_text))) # Count how many rows have empty or missing speech text
```

## Check totals and proportions in the summary table
```{r}
fear_prop <- tidy_data %>% # Use tidy_data dataset
  dplyr::filter(party %in% c("Labour", "Conservative")) %>% # Keep only rows where party is Labour or Conservative
  dplyr::group_by(year, party) %>% # Group the data by both year and party
  dplyr::summarise( # Create summary statistics 
    total_words = n(), # Count the total number of words in each group
    fear_words = sum(word %in% fear_dict), # Count how many words in each group are in the fear dictionary
    fear_prop = fear_words / total_words, # Calculate the proportion of fear words out of all words
    fear_per_1000 = (fear_words / total_words) * 1000, # Calculate the number of fear words per 1000 words
    .groups = "drop" # Remove grouping after summarising
  )

fear_prop # Display in summary table
```

## Combine a summary table of overall fear-word rates by party
```{r}
overall_fear_rate <- tidy_data %>% 
  dplyr::filter(party %in% c("Labour", "Conservative")) %>% 
  dplyr::group_by(party) %>% # Group the data by party
  dplyr::summarise( 
    total_words = dplyr::n(), 
    fear_words = sum(word %in% fear_dict), 
    fear_rate = fear_words / total_words, 
    fear_per_1000 = (fear_words / total_words) * 1000, 
    .groups = "drop" 
  )

overall_fear_rate # Display in summary table
```